Visualising How Non-linear Dimension Reduction Warps Your Data

### 2024-03-06 paper.qmd
Make the plan for the methodology

<!--In this paper, we introduce a novel approach, which is designed to determine which method, and which parameter choice provide the most useful representation of high dimensional data. Our approach involves dividing the data set into two subsets: a training set, which serves to construct the low-dimensional manifold, and a test set, utilized to generate predictive values and residuals.

For the implementation of our approach, a 2D embedding data set serves as the initiation point for the subsequent analysis and visualization (see @fig-nldervis (b)).-->

<!--### Overview {#sec-algorithm}-->

In this paper, we present a novel method that aims to determine the most effective way to represent high-D data by selecting the best method and hyperparameter choice. Our approach involves dividing the data set into two parts: a training set to construct the model and a test set to generate predictive values and residuals. To implement our approach, we first use a 2D embedding data set as the initial point. The output of our algorithm is a tour that displays the model and original data in the high-dimensional space. A flow chart of the proposed algorithm is shown in @fig-meth. Our algorithm comprises two main phases: (1) generate the model in the 2D space, and (2) generate the model in the high-D space. These two phases are described in detail in this section using UMAP 2D embedding of the S-curve data set, which has seven dimensions, including four noise dimensions.

![A flow diagram detailing the steps taken to create the low-dimensional manifold in the high dimensional space. There are two basic phases, one to generate the model in the 2D space, and other to generate the model in the high-D space.](figures/workflow.png){#fig-meth fig-align="center" width="100%" height="100%"}

@@

@@

```{r}
#| echo: false

@@

```{r}
#| echo: false

@@

```{r}
#| echo: false
#| warning: false
data <- read_rds("data/s_curve/s_curve.rds")
training_data <- read_rds("data/s_curve/s_curve_training.rds")
UMAP_data <- read_rds(file = "data/s_curve/s_curve_umap.rds")

@@

```{r}
#| warning: false
#| echo: false
#| label: fig-wkhighD
#| fig-cap: How the 2D model lift into high dimensions? (a) visualize the points and the hexagonal bin centroids related 2nd and 15th hexagons, (b) visualization of the edge connected the 2nd and 15th hexagons (colored in red) in the triangular mesh. A video of tour animation is available at <https://youtu.be/ffWFVK1UTIc>.

a2 + a1 +
  plot_annotation(tag_levels = 'a') +
  plot_layout(ncol = 3) &
  theme(legend.position='none', plot.tag = element_text(size = 8))
```

@@

```{r}
#| echo: false
#| message: false
#| warning: false
hex_full_count_df_loop1 <- read_rds("data/s_curve/s_curve_hex_3.rds")

@@

```{r}
#| echo: false
#| message: false
#| warning: false
hex_full_count_df_loop2 <- read_rds("data/s_curve/s_curve_hex_11.rds")

@@

```{r}
#| echo: false
#| message: false
#| warning: false
hex_full_count_df_loop3 <- read_rds("data/s_curve/s_curve_hex_20.rds")

@@

```{r}
#| echo: false
#| label: fig-binsize
#| fig-pos: H
#| fig-cap: "Hexbin plots from different number of bins for the **UMAP** projections of **S-curve** training data: (a) b = 32 (4, 8), s = 1.643542, (b) b = 264 (12, 22), s = 1.643542, and (c) b = 840 (21, 40), s = 1.643542. The hexbins are colored based on the density of points, with darker colors indicating higher point density and yellow color representing lower point density within each bin. What is the number of bins that would be effective in representing low-dimensional data?"

@@

@@

```{r}
#| echo: false
#| fig-cap: "Goodness of fit statistics from UMAP applied to training S-curve dataset. What is the effective number of bins in each NLDR technique to create a 2D model? The MSE plot have a steep slope at the beginning, indicating that a smaller number of bins causes a larger amount of error. Then, the slope gradually declines or level off, indicating that a higher number of bins generates a smaller error. Using the elbow method, when the total number of bins is set to 144, the slope of the Mean Squared Error (MSE) plot experiences a sudden and noticeable change, resembling an elbow-like shape. This point indicates that adding less bins does not enough to capture the data structure."
#| label: fig-diagnosticpltScurve
##| out-width: 100%
#| fig-pos: H

aic_plot + mse_plot +
  plot_annotation(tag_levels = 'a') +
  plot_layout(guides='collect', ncol = 2) &
  theme(legend.position='none', plot.tag = element_text(size = 8))
```

@@

```{r}
#| echo: false
#| message: false
#| warning: false

@@

```{r}
#| echo: false
#| warning: false
#| fig-cap: Goodness of f
#| label: fig-rmlowdenshex
#| fig-pos: H
##| out-width: 100%

@@

@@

<!--lwd smoothing criteria 1 with s-curve-->

@@

```{r}
#| echo: false
#| warning: false
#| fig-cap: "Goodness of fit statistics from UMAP applied to training S-curve dataset with different benchmark values to remove the low-density heaxgons. What is the effective benchmark value to remove the low-density heaxgons? The MSE plot have a steep slope at the beginning, indicating that a smaller benchmark causes a small amount of error. Then, the slope gradually increases or level up, indicating that a higher number of benchmark values generates a higher error. Using the reverse elbow method, when the benchmark value is set to 0.2, the slope of the Mean Squared Error (MSE) plot experiences a sudden and noticeable change, resembling an elbow-like shape. This point indicates that higher benchmark values remove the necessary bins as well which lead to distruct the 2D structure."
#| label: fig-diagnosticpltScurvelwd
##| out-width: 100%
#| fig-pos: H

aic_plot_lwd + mse_plot_lwd +
  plot_annotation(tag_levels = 'a') +
  plot_layout(guides='collect', ncol = 2) &
  theme(legend.position='none', plot.tag = element_text(size = 8))
```

@@

<!--The removal of long edges is a critical step to create a smoother representation  by iteratively eliminating hexagons with excessive distances between centroids. This process eliminates outliers and noise while preserving essential local relationships within the data. To achieve this, distances between vertices are sorted, and unique distance values are extracted. By setting a threshold based on the largest difference between consecutive distance values, long edges are identified and removed. This refinement step contributes to enhancing the quality of the triangular mesh, ensuring a more accurate representation of the data structure.-->

@@

```{r}
#| warning: false
#| echo: false
#| label: fig-modelScurverdistcount
#| fig-cap: "plots to identify benchmark values"
##| out-width: 100%
#| fig-pos: H

@@

```{r}
#| warning: false
#| echo: false
#| label: fig-modelScurvermlgimp
#| fig-cap: "What does the 2D model, constructed using UMAP (applied to the S-curve dataset), look like in high dimensions with long edges? (<https://youtu.be/pSL_G_n7-fM>) (a) 2D model without removing the long edges, and (b) 2D model with long edges coloured by red. The default benchmark value is used to identify long edges."
##| out-width: 100%
#| fig-pos: H

trimesh_umap_s_curve + trimesh_gr_umap_s_curve +
  #plot_annotation(tag_levels = 'a') +
  plot_layout(ncol = 2) &
  theme(legend.position='none', plot.tag = element_text(size = 8))
```

#### Hexagonal grid origin

@@

```{r}
#| echo: false
#| warning: false

@@

```{r}
#| echo: false
#| fig-cap: "(a) Visualization of bin centroids and (b) hexagonal grid, showcasing the configuration before (colored in gray) and after (colored in light orange) shifting the starting point by an identical amount applied in both x and y directions, with a shift value of $0.537285$."
#| label: fig-scurveshifthexgridsexp
##| out-width: 100%
#| fig-pos: H

@@

```{r}
#| echo: false
#| fig-cap: "Hexbin plots shows the distribution of points before and after shifting the starting point of the hexagonal grid generated for UMAP applied to the training S-curve dataset. (a) Hexbin plot before shifting, and (b) Hexbin plot after shifting (identical shift applied in both x and y directions, with a shift value of $0.537285$)."
#| label: fig-scurveshifthexgrids
##| out-width: 100%
#| fig-pos: H

@@

@@

```{r}
#| echo: false
#| fig-cap: "Goodness of fit statistics from UMAP applied to training S-curve dataset with different shift values. In this illustration, an identical shift is applied in both the x and y directions. The Mean Squared Error (MSE) plot exhibits a global minimum at $0$, indicating that no additional shifting is necessary for the current hexagonal grid to yield a well-fitted model."
#| label: fig-diagnosticpltScurvehexbins
##| out-width: 100%
#| fig-pos: H

aic_plot + mse_plot +
  plot_annotation(tag_levels = 'a') +
  plot_layout(guides='collect', ncol = 2) &
  theme(legend.position='none', plot.tag = element_text(size = 8))
```

@@

@@

```{r}
#| echo: false
#| fig-cap: "2D layouts from different NLDR techniques applied the same data: (a) tSNE (perplexity = 61), (b) UMAP (n_neighbors = 15), (c) PHATE (knn = 5), (d) TriMAP (n_inliers = 5, n_outliers = 4, n_random = 3), and (e) PaCMAP (n_neighbors = 10, init = random, MN_ratio = 0.9, FP_ratio = 2). Is there a best representation of the original data or are they all providing  equivalent information?"
#| label: fig-nldervis5Gau
#| out-width: 100%

@@

```{r}
#| warning: false
#| echo: false
#| label: fig-modelfiveGau
#| fig-cap: "Is there a best model to represent the original data in 2D space or are they all providing equivalent information?, (a) Model in the 2D space with tSNE (<https://youtu.be/x5VPB-wOLm4>), (b) Model in the 2D space with UMAP (<https://youtu.be/Cm1dW6iLyHQ>), (c) Model in the 2D space with PHATE (<https://youtu.be/HB0Y3ilz6sI>), (d) Model in the 2D space with TriMAP (<https://youtu.be/iSVkXIrOrSA>), and (e) Model in the 2D space with PaCMAP (<https://youtu.be/O2QBctsc4uM>)."
#| out-width: 100%

@@

```{r}
#| warning: false
#| echo: false
## PAHTE
## Prediction

@@

```{r}
#| echo: false
#| fig-cap: "Goodness of fit statistics from different NLDR techniques applied to training five spherical Gaussian cluster dataset. What is the best NLDR technique to represent the original data in 2D?"
#| label: fig-diagnosticpltGau
#| out-width: 100%

aic_gau_plot + mse_gau_plot +
  plot_annotation(tag_levels = 'a') +
  plot_layout(guides='collect', nrow = 2) &
  theme(legend.position='bottom', plot.tag = element_text(size = 8))
```

@@

duplicate_df_list <- mse_df_pbmc |>
  dplyr::filter(num_bins %in% dupli_bins) |>
  dplyr::arrange(num_bins) |>
  dplyr::group_split(num_bins)

duplicate_df <- data.frame(matrix(nrow = 0, ncol = 0))

@@

not_dupli_df <- mse_df_pbmc |>
  dplyr::filter(!(num_bins %in% dupli_bins))

mse_df_pbmc <- dplyr::bind_rows(not_dupli_df, duplicate_df)
```

@@

## To draw with AIC
aic_plot_pbmc <- ggplot(mse_df_pbmc, aes(x = num_bins, y = aic
)) +
  geom_point() +
  geom_line() +
  # geom_vline(xintercept = 256, linetype="dashed", 
  #               color = "red", size=0.5) +
  #geom_vline(xintercept = NROW(full_grid_with_hexbin_id)) +
  #annotate("text", x= (NROW(full_grid_with_hexbin_id) - 10), y=-5000, label=paste0("effective number of bins = ", as.character(NROW(full_grid_with_hexbin_id))), angle=90) +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  scale_colour_manual(values = c("#377eb8")) +
  ylab("AIC") +
  xlab("total number of bins")
## Effective number of bins along x-axis

mse_plot_pbmc <- ggplot(mse_df_pbmc, aes(x = num_bins,
                                       y = mse
)) +
  geom_point() +
  geom_line() +
   # geom_vline(xintercept = 256, linetype="dashed", 
   #              color = "red", size=0.5) +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  # geom_vline(xintercept = NROW(full_grid_with_hexbin_id)) +
  # annotate("text", x= (NROW(full_grid_with_hexbin_id) - 10), y=0.25, label=paste0("effective number of bins = ", as.character(NROW(full_grid_with_hexbin_id))), angle=90) +
  scale_colour_manual(values = c("#377eb8")) +
  ylab("MSE") +
  xlab("total number of bins")

@@

```{r}
#| warning: false
#| echo: false
# num_bins_pbmc <- calculate_effective_x_bins(.data = UMAP_pbmc, x = UMAP1,
#                                               cell_area = 1) ##23

num_bins_pbmc <- 15

## To extract bin centroids
hexbin_data_object_pbmc <- extract_hexbin_centroids(nldr_df = UMAP_pbmc,
                                                      num_bins = num_bins_pbmc,
                                                      shape_val = shape_value_pbmc, x = UMAP1, y = UMAP2)

df_bin_centroids_pbmc <- hexbin_data_object_pbmc$hexdf_data

#num_non_empty_bins_pbmc <- df_bin_centroids_pbmc$hexID |> length() #83

UMAP_pbmc_with_hb_id <- UMAP_pbmc |>
  dplyr::mutate(hb_id = hexbin_data_object_pbmc$hb_data@cID)

## To generate a data set with high-D and 2D training data
df_all_pbmc <- dplyr::bind_cols(training_data_pbmc |> dplyr::select(-ID), UMAP_pbmc_with_hb_id)

## Averaged on high-D
df_bin_pbmc <- avg_highD_data(.data = df_all_pbmc, column_start_text = "PC")

## Triangulate bin centroids
tr1_object_pbmc <- triangulate_bin_centroids(df_bin_centroids_pbmc, x, y)
tr_from_to_df_pbmc <- generate_edge_info(triangular_object = tr1_object_pbmc)

## Compute 2D distances
distance_pbmc <- cal_2d_dist(.data = tr_from_to_df_pbmc)

distance_plot <- plot_dist(distance_pbmc) +
  #ggtitle("(b)" ) +
  ylab(expression(d^{(2)})) +
  theme(axis.text = element_text(size = 5),
        axis.title = element_text(size = 12))

@@

## To find the benchmark value
benchmark_pbmc <- find_benchmark_value(.data = distance_pbmc, distance_col = distance)
benchmark_pbmc <- 2.5

trimesh_removed_pbmc_umap <- remove_long_edges(.data = distance_pbmc, benchmark_value = benchmark_pbmc,
                                                 triangular_object = tr1_object_pbmc, distance_col = distance)

trimesh_removed_pbmc_umap <- trimesh_removed_pbmc_umap +
  geom_point(size = 1, colour = "#33a02c") +
  #ggtitle("(b)") +
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'b', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

@@

@@

error_df <- error_df |>
  mutate(type = if_else(total <= 2, "error 2 or less",
                        if_else(total <= 10, "error 2-10",
                                if_else(total <= 15, "error 10-15",
                                        if_else(total <= 20, "error 15-20",
                                                if_else(total <= 25, "error 20-25",
                                                        if_else(total <= 30, "error 25-30",
                                                                if_else(total <= 35, "error 30-35", "error greter than 35"))))))))

plot_list2_pbmc_error <- error_df |>
  mutate(type = factor(type , levels = c("error 2 or less", "error 2-10", "error 10-15", "error 15-20", "error 20-25", "error 25-30", "error 30-35", "error greter than 35"))) |>
    ggplot(aes(x = UMAP1,
               y = UMAP2, color = type))+
    geom_point(alpha=0.5, size = 0.1) +
    coord_equal() +
    theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold")) + #ggtitle("(a)") +
  theme_linedraw() +
    theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm')) +
  scale_color_manual(values=c("#fee0d2", "#fcbba1",
                                         "#fc9272", "#fb6a4a", "#ef3b2c",
                                         "#cb181d", "#a50f15", "#99000d")) 
```

>>>>

In this paper, we introduce a novel method to determine the most effective NLDR technique and the best hyperparameter choice that provides the most useful representation of high-D data. Our approach involves dividing the high-dimensional dataset into two parts: a training set for constructing the model and a test set for generating predictive values and residuals. Our algorithm takes a 2D embedding data as the input and generate a tour that displays the high-dimensional wireframe to overlay the data. The flow chart of the proposed algorithm is shown in @fig-meth. The algorithm consists of two main phases: (1) generating the model in the 2D space and (2) lifting the model into high-D space. The main steps of the algorithm are described in detail in this section using UMAP 2D embedding of the S-curve dataset. This dataset has seven dimensions, including four noise dimensions that were added to the original 3D data.

![The flow diagram shows the main steps of our algorithm. There are two basic phases, one to generate the model in the 2D space, and other to map the model into the high-D space.](figures/workflow.png){#fig-meth fig-align="center" width="100%" height="100%"}

@@

**Step 1: Scaling NLDR data**

**Step 2: Hexagonating NLDR data**

**(b) Determine the number of bins along the y axis** ($b_2$)

**(d) Assign NLDR data to hexagons**

**Step 3: Obtaining bin centroids or bin means**

**Step 4: Triangulating bin centroids or bin means**

@@

<!--
```{r}
#| echo: false
#| eval: false

@@

```{r}
#| echo: false
#| eval: false

@@

```{r}
#| echo: false
#| warning: false
#| eval: false

data <- read_rds("data/s_curve/s_curve.rds")
training_data <- read_rds("data/s_curve/s_curve_training.rds")
UMAP_data <- read_rds(file = "data/s_curve/s_curve_umap.rds")

@@

```{r}
#| eval: false
#| warning: false
#| echo: false
#| label: fig-wkhighD
#| fig-cap: How the 2D model lift into high dimensions? (a) visualize the points and the hexagonal bin centroids related 2nd and 15th hexagons, (b) visualization of the edge connected the 2nd and 15th hexagons (colored in red) in the triangular mesh. A video of tour animation is available at <https://youtu.be/ffWFVK1UTIc>.

a2 + a1 +
  plot_annotation(tag_levels = 'a') +
  plot_layout(ncol = 3) &
  theme(legend.position='none', plot.tag = element_text(size = 8))
```
-->

@@

<!--                  
```{r}
#| echo: false
#| message: false
#| warning: false
#| eval: false
hex_full_count_df_loop1 <- read_rds("data/s_curve/s_curve_hex_3.rds")

@@

```{r}
#| echo: false
#| message: false
#| warning: false
#| eval: false
hex_full_count_df_loop2 <- read_rds("data/s_curve/s_curve_hex_11.rds")

@@

```{r}
#| echo: false
#| message: false
#| warning: false
#| eval: false
hex_full_count_df_loop3 <- read_rds("data/s_curve/s_curve_hex_20.rds")

@@

```{r}
#| eval: false
#| echo: false
#| label: fig-binsize
#| fig-pos: H
#| fig-cap: "Hexbin plots from different number of bins for the **UMAP** projections of **S-curve** training data: (a) b = 32 (4, 8), s = 1.643542, (b) b = 264 (12, 22), s = 1.643542, and (c) b = 840 (21, 40), s = 1.643542. The hexbins are colored based on the density of points, with darker colors indicating higher point density and yellow color representing lower point density within each bin. What is the number of bins that would be effective in representing low-dimensional data?"

@@

```{r}
#| warning: false
#| echo: false
#| eval: false

@@

```{r}
#| eval: false
#| echo: false
#| fig-cap: "Goodness of fit statistics from UMAP applied to training S-curve dataset. What is the effective number of bins in each NLDR technique to create a 2D model? The MSE plot have a steep slope at the beginning, indicating that a smaller number of bins causes a larger amount of error. Then, the slope gradually declines or level off, indicating that a higher number of bins generates a smaller error. Using the elbow method, when the total number of bins is set to 144, the slope of the Mean Squared Error (MSE) plot experiences a sudden and noticeable change, resembling an elbow-like shape. This point indicates that adding less bins does not enough to capture the data structure."
#| label: fig-diagnosticpltScurve
##| out-width: 100%
#| fig-pos: H

aic_plot + mse_plot +
  plot_annotation(tag_levels = 'a') +
  plot_layout(guides='collect', ncol = 2) &
  theme(legend.position='none', plot.tag = element_text(size = 8))
```
-->

@@

<!--
```{r}
#| echo: false
#| message: false
#| warning: false
#| eval: false

@@

```{r}
#| eval: false
#| echo: false
#| warning: false
#| fig-cap: Goodness of f
#| label: fig-rmlowdenshex
#| fig-pos: H
##| out-width: 100%

@@

-->

@@

<!--lwd smoothing criteria 1 with s-curve-->
<!--
```{r}
#| warning: false
#| echo: false
#| eval: false

@@

```{r}
#| eval: false
#| echo: false
#| warning: false
#| fig-cap: "Goodness of fit statistics from UMAP applied to training S-curve dataset with different benchmark values to remove the low-density heaxgons. What is the effective benchmark value to remove the low-density heaxgons? The MSE plot have a steep slope at the beginning, indicating that a smaller benchmark causes a small amount of error. Then, the slope gradually increases or level up, indicating that a higher number of benchmark values generates a higher error. Using the reverse elbow method, when the benchmark value is set to 0.2, the slope of the Mean Squared Error (MSE) plot experiences a sudden and noticeable change, resembling an elbow-like shape. This point indicates that higher benchmark values remove the necessary bins as well which lead to distruct the 2D structure."
#| label: fig-diagnosticpltScurvelwd
##| out-width: 100%
#| fig-pos: H

aic_plot_lwd + mse_plot_lwd +
  plot_annotation(tag_levels = 'a') +
  plot_layout(guides='collect', ncol = 2) &
  theme(legend.position='none', plot.tag = element_text(size = 8))
```
-->

@@

<!--The removal of long edges is a critical step to create a smoother representation  by iteratively eliminating hexagons with excessive distances between centroids. This process eliminates outliers and noise while preserving essential local relationships within the data. To achieve this, distances between vertices are sorted, and unique distance values are extracted. By setting a threshold based on the largest difference between consecutive distance values, long edges are identified and removed. This refinement step contributes to enhancing the quality of the triangular mesh, ensuring a more accurate representation of the data structure.-->
<!--
```{r}
#| warning: false
#| echo: false
#| eval: false

@@

```{r}
#| eval: false
#| warning: false
#| echo: false
#| label: fig-modelScurverdistcount
#| fig-cap: "plots to identify benchmark values"
##| out-width: 100%
#| fig-pos: H

@@

```{r}
#| eval: false
#| warning: false
#| echo: false
#| label: fig-modelScurvermlgimp
#| fig-cap: "What does the 2D model, constructed using UMAP (applied to the S-curve dataset), look like in high dimensions with long edges? (<https://youtu.be/pSL_G_n7-fM>) (a) 2D model without removing the long edges, and (b) 2D model with long edges coloured by red. The default benchmark value is used to identify long edges."
##| out-width: 100%
#| fig-pos: H

trimesh_umap_s_curve + trimesh_gr_umap_s_curve +
  #plot_annotation(tag_levels = 'a') +
  plot_layout(ncol = 2) &
  theme(legend.position='none', plot.tag = element_text(size = 8))
```
-->

#### Starting point of the hexagonal grid

@@

<!--

```{r}
#| eval: false
#| echo: false
#| warning: false

@@

```{r}
#| eval: false
#| echo: false
#| fig-cap: "(a) Visualization of bin centroids and (b) hexagonal grid, showcasing the configuration before (colored in gray) and after (colored in light orange) shifting the starting point by an identical amount applied in both x and y directions, with a shift value of $0.537285$."
#| label: fig-scurveshifthexgridsexp
##| out-width: 100%
#| fig-pos: H

@@

```{r}
#| eval: false
#| echo: false
#| fig-cap: "Hexbin plots shows the distribution of points before and after shifting the starting point of the hexagonal grid generated for UMAP applied to the training S-curve dataset. (a) Hexbin plot before shifting, and (b) Hexbin plot after shifting (identical shift applied in both x and y directions, with a shift value of $0.537285$)."
#| label: fig-scurveshifthexgrids
##| out-width: 100%
#| fig-pos: H

@@

```{r}
#| eval: false
#| warning: false
#| echo: false

@@

```{r}
#| eval: false
#| echo: false
#| fig-cap: "Goodness of fit statistics from UMAP applied to training S-curve dataset with different shift values. In this illustration, an identical shift is applied in both the x and y directions. The Mean Squared Error (MSE) plot exhibits a global minimum at $0$, indicating that no additional shifting is necessary for the current hexagonal grid to yield a well-fitted model."
#| label: fig-diagnosticpltScurvehexbins
##| out-width: 100%
#| fig-pos: H

aic_plot + mse_plot +
  plot_annotation(tag_levels = 'a') +
  plot_layout(guides='collect', ncol = 2) &
  theme(legend.position='none', plot.tag = element_text(size = 8))
```
-->

@@

<!--
```{r}
#| warning: false
#| echo: false
#| eval: false

@@

```{r}
#| eval: false
#| echo: false
#| fig-cap: "2D layouts from different NLDR techniques applied the same data: (a) tSNE (perplexity = 61), (b) UMAP (n_neighbors = 15), (c) PHATE (knn = 5), (d) TriMAP (n_inliers = 5, n_outliers = 4, n_random = 3), and (e) PaCMAP (n_neighbors = 10, init = random, MN_ratio = 0.9, FP_ratio = 2). Is there a best representation of the original data or are they all providing  equivalent information?"
#| label: fig-nldervis5Gau
#| out-width: 100%

@@

```{r}
#| eval: false
#| warning: false
#| echo: false
#| label: fig-modelfiveGau
#| fig-cap: "Is there a best model to represent the original data in 2D space or are they all providing equivalent information?, (a) Model in the 2D space with tSNE (<https://youtu.be/x5VPB-wOLm4>), (b) Model in the 2D space with UMAP (<https://youtu.be/Cm1dW6iLyHQ>), (c) Model in the 2D space with PHATE (<https://youtu.be/HB0Y3ilz6sI>), (d) Model in the 2D space with TriMAP (<https://youtu.be/iSVkXIrOrSA>), and (e) Model in the 2D space with PaCMAP (<https://youtu.be/O2QBctsc4uM>)."
#| out-width: 100%

@@

```{r}
#| eval: false
#| warning: false
#| echo: false
## PAHTE
## Prediction

@@

```{r}
#| eval: false
#| echo: false
#| fig-cap: "Goodness of fit statistics from different NLDR techniques applied to training five spherical Gaussian cluster dataset. What is the best NLDR technique to represent the original data in 2D?"
#| label: fig-diagnosticpltGau
#| out-width: 100%

aic_gau_plot + mse_gau_plot +
  plot_annotation(tag_levels = 'a') +
  plot_layout(guides='collect', nrow = 2) &
  theme(legend.position='bottom', plot.tag = element_text(size = 8))
```
-->

@@

### Group split by duplicated bins
duplicate_df_list <- mse_df_pbmc |>
  dplyr::filter(num_bins %in% dupli_bins) |>
  dplyr::arrange(num_bins) |>
  dplyr::group_split(num_bins)

### Obtain one row from duplicates which have lowest error and hexsize
duplicate_df <- data.frame(matrix(nrow = 0, ncol = 0))

@@

### Obtain the mse_df with not duplicated bins
not_dupli_df <- mse_df_pbmc |>
  dplyr::filter(!(num_bins %in% dupli_bins))

### Combine duplicated and not duplicated(corrected) bins dfs
mse_df_pbmc <- dplyr::bind_rows(not_dupli_df, duplicate_df)
```

@@

## To draw with AIC
aic_plot_pbmc <- ggplot(mse_df_pbmc, aes(x = num_bins, y = aic
)) +
  geom_point() +
  geom_line() +
  geom_vline(xintercept = 80, linetype="solid",
                color = "red", size=0.8, alpha = 0.5) +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  #scale_colour_manual(values = c("#377eb8")) +
  ylab("AIC") +
  xlab("total number of bins")
## Effective number of bins along x-axis

mse_plot_pbmc <- ggplot(mse_df_pbmc, aes(x = num_bins,
                                       y = mse
)) +
  geom_point() +
  geom_line() +
   geom_vline(xintercept = 80, linetype="solid",
                color = "red", size=0.8, alpha = 0.5) +
  theme_light() +
  theme(legend.position = "none", legend.title = element_blank(), plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title = element_text(size = 7),
        axis.text = element_text(size = 7)) +
  #scale_colour_manual(values = c("#377eb8")) +
  ylab("MSE") +
  xlab("total number of bins")

@@

## Decide by looking at MSE plot
num_bins_x_pbmc <- 8
num_bins_y_pbmc <- 10
hex_size_pbmc <- 0.1

hb_obj_pbmc <- hex_binning(data = umap_pbmc_scaled |> dplyr::select(-cell_label), x = "UMAP1", 
                      y = "UMAP2", num_bins_x = num_bins_x_pbmc, 
                      num_bins_y = num_bins_y_pbmc, x_start = NA, y_start = NA, 
                      buffer_x = NA, buffer_y = NA, hex_size = hex_size_pbmc, col_start = "UMAP")

## Data set with all possible centroids in the hexagonal grid
all_centroids_df <- as.data.frame(do.call(cbind, hb_obj_pbmc$centroids))

## Generate all coordinates of hexagons
hex_grid <- as.data.frame(do.call(cbind, hb_obj_pbmc$hex_poly))

## To obtain the standardise counts within hexbins
counts_df <- as.data.frame(do.call(cbind, hb_obj_pbmc$std_cts))
df_bin_centroids <- extract_hexbin_centroids(centroids_df = all_centroids_df, 
                                             counts_df = counts_df)
ggplot(data = hex_grid, aes(x = x, y = y)) + 
  geom_polygon(fill = "white", color = "black", aes(group = hex_poly_id)) +
  geom_point(data = all_centroids_df, aes(x = c_x, y = c_y), color = "red") +
  coord_fixed()

hex_grid_with_counts <- dplyr::left_join(hex_grid, counts_df, by = c("hex_poly_id" = "hb_id"))

ggplot(data = hex_grid_with_counts, aes(x = x, y = y)) +
  geom_polygon(color = "black", aes(group = hex_poly_id, fill = std_counts)) +
  geom_text(data = all_centroids_df, aes(x = c_x, y = c_y, label = hexID)) +
  scale_fill_viridis_c(direction = -1, na.value = "#ffffff") +
  coord_fixed()

umap_data_with_hb_id_pbmc <- as.data.frame(do.call(cbind, hb_obj_pbmc$data_hb_id))

model_object <- fit_highd_model( training_data = training_data_pbmc, 
                                 nldr_df_with_id = umap_pbmc_scaled |> dplyr::select(-cell_label), 
                                 x = "UMAP1", y = "UMAP2", 
                                 num_bins_x = num_bins_x_pbmc, 
                                 num_bins_y = num_bins_y_pbmc, 
                                 x_start = NA, y_start = NA, 
                                 buffer_x = NA, buffer_y = NA, 
                                 hex_size = hex_size_pbmc,
                                 is_rm_lwd_hex = FALSE, 
                                 benchmark_to_rm_lwd_hex = NA, 
                                 col_start_2d = "UMAP", 
                                 col_start_highd = "PC_")

df_bin_centroids_pbmc <- model_object$df_bin_centroids
df_bin_pbmc <- model_object$df_bin

## Triangulate bin centroids
tr1_object_pbmc <- tri_bin_centroids(df_bin_centroids_pbmc, x = "c_x", y = "c_y")
tr_from_to_df_pbmc <- gen_edges(tri_object = tr1_object_pbmc)

## Compute 2D distances
distance_pbmc <- cal_2d_dist(tr_coord_df = tr_from_to_df_pbmc, 
                             start_x = "x_from", start_y = "y_from", 
                             end_x = "x_to", end_y = "y_to", 
                             select_vars = c("from", "to", "distance"))

distance_plot <- plot_dist(distance_pbmc) +
  ylab(expression(d^{(2)})) +
  theme(axis.text = element_text(size = 5),
        axis.title = element_text(size = 12))

@@

## To find the benchmark value
benchmark_pbmc <- find_lg_benchmark(distance_edges = distance_pbmc, distance_col = "distance")

ggplot() +
geom_trimesh(data = df_bin_centroids_pbmc, mapping = aes(x = c_x, y = c_y))

trimesh_removed_pbmc_umap <- vis_rmlg_mesh(distance_edges = distance_pbmc, benchmark_value = benchmark_pbmc, tr_coord_df = tr_from_to_df_pbmc, distance_col = "distance")

trimesh_removed_pbmc_umap <- trimesh_removed_pbmc_umap +
  theme_linedraw() +
  theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
        axis.title.x = element_blank(), axis.title.y = element_blank(),
        axis.text.x = element_blank(), axis.ticks.x = element_blank(),
        axis.text.y = element_blank(), axis.ticks.y = element_blank(),
        panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  annotate(geom = 'text', label = 'b', x = -Inf, y = Inf, hjust = -0.5, vjust = 1.5, size = 3)

@@

df_all_pbmc <- dplyr::bind_cols(training_data_pbmc |> dplyr::select(-ID), umap_data_with_hb_id_pbmc)

@@

error_df <- error_df |>
  mutate(type = if_else(total == 0, "no error",
                        if_else(total <= 15, "error 1-15",
                                if_else(total <= 20, "error 15-20",
                                        if_else(total <= 25, "error 20-25",
                                                if_else(total <= 30, "error 25-30",
                                                        if_else(total <= 35, "error 30-35",
                                                                if_else(total <= 40, "error 35-40", "error greter than 40"))))))))

plot_list2_pbmc_error <- error_df |>
  mutate(type = factor(type , levels = c("no error", "error 1-15", "error 15-20", "error 20-25", "error 25-30", "error 30-35", "error 35-40", "error greter than 40"))) |>
    ggplot(aes(x = UMAP1,
               y = UMAP2, color = type))+
    geom_point(alpha=0.5, size = 0.1) +
    coord_equal() +
    theme(plot.title = element_text(hjust = 0.5, size = 18, face = "bold")) + #ggtitle("(a)") +
  theme_linedraw() +
    theme(plot.title = element_text(size = 7, hjust = 0.5, vjust = -0.5),
              axis.title.x = element_blank(), axis.title.y = element_blank(),
              axis.text.x = element_blank(), axis.ticks.x = element_blank(),
              axis.text.y = element_blank(), axis.ticks.y = element_blank(),
              panel.grid.major = element_blank(), panel.grid.minor = element_blank(), #change legend key width
        legend.title = element_text(size=5), #change legend title font size
        legend.text = element_text(size=4),
         legend.key.height = unit(0.25, 'cm'),
         legend.key.width = unit(0.25, 'cm')) +
  scale_color_manual(values=c("#fee0d2", "#fcbba1",
                                         "#fc9272", "#fb6a4a", "#ef3b2c",
                                         "#cb181d", "#a50f15", "#99000d")) 
```
